---
title: "The education-obesity link across countries over time"
author: "Anastasios Goudras(2855565), Thedor(), Stefan(), Omar(), Ingmar(2890403), Diego Melendez(2841772)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>")
library(tidyverse)
```

# Title Page

Tutorial group: 6.1

Tutorial lecturer: Simon Loop

# Part 1 - Identify a Social Problem

The social problem we have identified is the link between the BMI Index and educational attainment level, bridging two data sets from Eurostat to compare and identify any correlation between the two factors. 

## 1.1 Describe the Social Problem

The relevance of these two variables, BMI and education levels, and how they are correlated with each other lies in the fact that obesity comes with serious health risks, such as diabetes and cardiovascular issues, which are often more common among people with lower educational attainment (World Health Organization, 2022; European Commission, 2024). This then results in the burden of health risks being disproportionately put on the lower educated class of society, and further increases social inequality issues. Beyond the health risks themselves, education has a clear impact on the ability of a person to maintain a healthy lifestyle, as higher education often comes with higher income, better access to higher quality food and so on. This means that addressing obesity cannot be done through increasing awareness alone, and that broader structural interventions, such as food pricing policies or workplace conditions need to be considered.

The fact we are looking at the European aspect of this issue also adds to its relevance. By looking at multiple countries over time, it becomes possible to assess whether this education gap in obesity rates is a universal pattern across Europe or whether certain countries have managed to narrow it. This kind of cross country comparison is useful for understanding what policy approaches may actually work in practice, especially since educational inequalities in obesity have been documented across Europe in recent research (World Health Organization, 2022; European Commission, 2024). 

While the existence of an educational gradient in obesity is well established in the literature (World Health Organization, 2022; European Commission, 2024), far less is settled about how this gap behaves over time and whether it is narrowing or widening across different European countries. Much existing work documents the association at a single point in time or within individual countries, which leaves the question of long-term, cross country trends comparatively underexplored. This is the gap our project addresses: by tracking obesity rates by education level across multiple European countries and survey years, we aim to establish whether the education obesity divide is a stable, universal pattern or one that some countries have managed to reduce.


# Part 2 - Data Sourcing

## 2.1 Load in the data

Both datasets are published by Eurostat. We downloaded them once and saved them as CSV files in the `data/` folder of our repository, which we read into R with `read_csv()`. Using saved copies rather than querying Eurostat live means the report always knits from the same data and does not depend on a network connection or on the Eurostat servers being reachable. The files are small, so storing them in the repository keeps the project self-contained and reproducible: anyone who clones it can knit the report without extra setup.

```{r loading_data}

# Load the saved datasets from our repository
bmi_edu <- read_csv("data/bmi_edu.csv")
obesity_trend <- read_csv("data/obesity_trend.csv")

```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(bmi_edu)
head(obesity_trend)
```
Our first dataset, bmi_edu (Eurostat code hlth_ehis_bm1e, Body mass index (BMI) by sex, age and educational attainment level), comes from the European Health Interview Survey (EHIS), in which respondents report their own height and weight (Eurostat, n.d.-a). For each country and year it gives the share of people in each BMI category, broken down by education level, sex, and age. EHIS runs only every few years, so the data covers mainly the 2014 and 2019 waves. This is the dataset that links weight to education, so it is central to our analysis.

Our second dataset, obesity_trend (Eurostat code sdg_02_10, Obesity rate by body mass index (BMI)), is one of the EU's Sustainable Development Goal indicators, also based on self-reported health surveys (Eurostat, n.d.-b). It does not split the figures by education, but it covers many more years (2008, 2014, 2017, 2019, 2022), which makes it well suited to tracking obesity over time. It supplies the "across countries, over time" side of our research question. The data are reported by country, covering the EU member states together with several additional European countries, each shown as a two-letter code in the geo column.

## 2.3 Describe the type of variables included

```{r variables_table, echo=FALSE}
variables <- tribble(
  ~Variable,     ~Type,        ~Dataset,   ~Meaning,
  "freq",        "character",  "both",     "data frequency (A = annual)",
  "unit",        "character",  "both",     "unit of measurement (PC = percentage)",
  "bmi",         "character",  "both",     "BMI category (e.g. overweight, obese)",
  "isced11",     "character",  "bmi_edu",  "education level (ISCED 2011 group)",
  "sex",         "character",  "bmi_edu",  "sex (M, F, T = total)",
  "age",         "character",  "bmi_edu",  "age group",
  "geo",         "character",  "both",     "country (Eurostat code)",
  "TIME_PERIOD", "numeric",    "both",     "survey year",
  "values",      "numeric",    "both",     "share of people in that category"
)

knitr::kable(variables, caption = "Variables in the two datasets.")
```

As Table 1 shows, the variables fall into three groups. The health measure is BMI/obesity itself, the share of people in each weight category (the `bmi` and `values` columns). The socioeconomic measure is education (the `isced11` column in `bmi_edu`), which is the variable at the heart of our question. The remaining columns (`sex`, `age`, `geo` for country, and `TIME_PERIOD` for year) are background characteristics that we use to split the figures by group, place, and time.

Neither dataset comes from administrative records; both rely on surveys in which people report their own height and weight. This has two implications. First, self reported data tends to underestimate obesity slightly, since people often round their weight down. Second, the two datasets are not fully independent: the SDG indicator is partly built on the same EHIS survey as `bmi_edu`, so they are related rather than separate sources. We combined them deliberately: `bmi_edu` provides the education breakdown, while `obesity_trend` adds the longer time span. Because both are self reported survey data, our analysis can describe associations (for example, that obesity is more common among lower educated groups) but cannot show that education causes obesity.


# Part 3 - Quantifying

## 3.1 Data cleaning

Stefan add explanation here 

**Cleaning Step 1: Filter and clean the first dataset**
```{r}
bmi_edu_clean <- bmi_edu %>%
  filter(
    bmi == "BMI_GE30",                        # Keep only Obese rows
    isced11 %in% c("ED0-2", "ED3_4", "ED5-8"), # Separate the 3 education groups
    sex == "T",                              # Combined sexes
    age == "TOTAL"                          # Combined adult age group
  ) %>%
  select(geo, TIME_PERIOD, isced11, bmi_value = values) %>%
  drop_na(bmi_value)
  
```

and here 

**Cleaning Step 2: Filter and clean the second dataset**
```{r}

obesity_trend_clean <- obesity_trend %>%
  filter(
    bmi == "BMI_GE30" # Keep only Obese rows since that's what we are comparing
  ) %>%
  select(geo, TIME_PERIOD, trend_value = values) %>%
  drop_na(trend_value)
```

and here 
  
**Cleaning Step 3: Combine both datasets into one master file**
```{r}
combined_dataset <- inner_join(bmi_edu_clean, obesity_trend_clean, by = c("geo", "TIME_PERIOD"))
```

and here 

## 3.2 Generate necessary variables

and here 


Variable 1

```{r gen_var1}
# Create Variable 1: Human-readable education levels
bmi_edu_with_var1 <- combined_dataset %>%
  mutate(
    edu_level = case_when(
      isced11 == "ED0-2" ~ "Low Education",
      isced11 == "ED3_4" ~ "Medium Education",
      isced11 == "ED5-8" ~ "High Education"
    )
  )
```


```{r}
# Reshaping before subtracting
gap_data <- bmi_edu_with_var1 %>%
  # 1. Isolate the columns we need
  select(geo, TIME_PERIOD, edu_level, bmi_value) %>%
  
  # 2.Flip the rows into columns
  pivot_wider(
    names_from = edu_level, 
    values_from = bmi_value
  )
```


Variable 2

```{r gen_var2}

# Create Variable 2: Calculate the obesity inequality gap
final_gap_data <- gap_data %>%
  mutate(
    obesity_gap = `Low Education` - `High Education`
  )
```

## 3.3 Visualize temporal variation

```{r}
# First I make a vector of the landcodes of the countries in Europe with the largest population, because using every country from the datasets will make the visualisation really chaotic.
largest_european_countries <- c("RU", "TR", "DE", "UK", "FR", "IT", "ES", "UA", "PL", "RO")
# RU:Russia, TR:Turkey, DE:Germany, UK:UnitedKingdom, FR:France, IT:Italy, ES:Spain, UA:Ukraine, PL:Poland, RO:Romania

# The following code only selects the countries that are ACTUALLY in the dataset. So now if Russia, for example, is not in the dataset, R does not crash.
selected_countries <- largest_european_countries[
  largest_european_countries %in% obesity_trend_clean$geo
]

# This line shows the countries that are selected.
selected_countries

obesity_trend_clean %>% 
  filter(geo %in% selected_countries) %>% 
  ggplot(aes(x = TIME_PERIOD, y = trend_value, color = geo, group = geo)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
    labs(
      title = "Temporal variation in obesity rates",
      subtitle = "Largest European countries in the dataset",
      x = "Year",
      y = "Obesity rate (%)",
      color = "Country"
    ) +
    theme_minimal()
# line 185: we can not use final_gap_data because otherwise we do not have the required periods since this only has 2014 and 2019, that is why I used obesity_trend_clean because now we have more than enough periods.
# Line 186: This filters only the selected countries from the dataset
# Line 187: this plots the obesity over time, x is the years on the x-axis, y is the obesity rate on the y-axis, color equals geo gives every country a different color, and group=geo connects all the lines.
# Lines 188-197: the codes show how the lines are shown for example the width etc, it also creates the titles and the info of everything that is shown in the visualisation.
  
```

## 3.4 Visualize spatial variation

```{r visualise_map}
latest_year <- max(final_gap_data$TIME_PERIOD, na.rm = TRUE)

final_gap_data %>%
  filter(TIME_PERIOD == latest_year) %>%
  ggplot(aes(x = reorder(geo, obesity_gap), y = obesity_gap)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste("Spatial variation in the education-obesity gap in", latest_year),
    subtitle = "Difference between low and high education obesity rates across countries",
    x = "Country",
    y = "Obesity gap, percentage points"
  ) +
  theme_minimal()

# Line 208 selects the last year that we are able to use
# Line 211 filters out everything except the last year
# Line 212-221: creates the plot geom_col makes the bar plot, coord_flip flips the plot so it is nicer to read.The last few lines create the titles and labels.
```

Here you provide a description of why the plot above is relevant to your specific social problem.

This figure shows the regional differences in the education-obesity gap between countries in the most recent year for which data is available. The gap is calculated by subtracting the percentage of obese people among the low-skilled from the percentage of obese people among the highly skilled.

A positive value indicates that obesity is more prevalent among the low-educated. The countries are ranked according to the size of the gap, making it easier to see where education-related inequality in obesity is greatest.

## 3.5 Visualize sub-population variation

What is the obesity rate by country?

#```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables (BMI .

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

https://github.com/theodorkubi/ProgrammingForEcon

## 5.2 Reference list

European Commission. (2024). Health inequalities: Overweight and obesity. Knowledge4Policy. https://knowledge4policy.ec.europa.eu/health-promotion-knowledge-gateway/health-inequalities-5_en

World Health Organization. (2022). Obesity and inequities. https://iris.who.int/bitstreams/c025afdb-1ca7-49e2-ba89-adca1529e47c/download

Eurostat. (n.d.-a). Body mass index (BMI) by sex, age and educational attainment level [Data set]. Retrieved June 7, 2026, from https://ec.europa.eu/eurostat/databrowser/view/hlth_ehis_bm1e

Eurostat. (n.d.-b). Obesity rate by body mass index (BMI) [Data set]. Retrieved June 7, 2026, from https://ec.europa.eu/eurostat/databrowser/view/sdg_02_10